---
title: "607 Assignment 5A Dylan Gold"
format: html
editor: visual
---

# 607 Assignment 5A Dylan Gold

## Approach

In this assignment we need to recreate a table given to us from an image, then tidy that data set.\
When tidying a data set there are several things we have to do. Specifically we need to recreate the data in a way such that each column is a variable, row is a observation and each cell is a value. Then we need to count data and some sort of analysis with this, compare percentages and other things shown in the rubric.

My plan is to recreate the data in csv format. I will just do it by hand because it is only 20 cell values and afew row and column names.\
I will add an extra row name, which is implied in the data but it would be easier to work with if it already had the row value.\
Bring it into R and use tidyr and dplyr with to make it molten/long and go from there to perform the analysis. Functions like pivot_longer and pivot_wider will help.

# Codebase
Setup libraries
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
```


First I manually make a csv from the image, then I import it in after pushing it to github.



```{r,message=FALSE, warning=FALSE}
url <- "https://raw.githubusercontent.com/DylanGoldJ/607-Assignment-4/refs/heads/main/5A_Data.csv"

df <- read_csv(
  file = url,
  col_names = TRUE
)

df
```

We can start to make it tidy. We can start by fixing the first two column names
```{r}
airport_arrivals <- df %>% rename(c(airline = "...1", arrival_status = "...2"))
airport_arrivals
```
Now we can make it tidy by combining all the columns into row values. We need to have a row specifying the destination, as well as a column containing the frequency displayed in the cells in the data frame above.
```{r}
airport_arrivals <- airport_arrivals %>%
  pivot_longer(
    cols = !c(airline:arrival_status),
    names_to = "destination",
    values_to = "frequency"
    )

head(airport_arrivals, 8)

```

The data is now in tidy form. We can now start to analys #TODO








